Interpreting the gas-fraction as the entropy of an isothermal dusty gas

### 2016-07-30 linear.tex
adding results

\section{Linear problem} 
We consider radially-localized axisymmetric disturbances of the form 
\begin{align}
  \delta\rho(r, z)\exp{\left(\ii k r-\ii\sigma t\right)},
\end{align} 
and similarly for $\dd P$ and $\dd\bm{v}$. Here, $k$ is a real
wavenumber such that $|kr|\gg 1$, and the amplitude $\dd\rho(r,z)$ is
a slowly-varying function of $r$. Then 
$\p_r\to i k$ when acting on the above primitive perturbations, and we may
neglect curvature terms. 
We take  
$k>0$ without loss of  generality. The frequency $\sigma = \omega +
\ii s$ is generally complex, with $\omega$ being the real frequency
and $s$ is the real growth rate.

Introducing 
\begin{align}
  W \equiv \frac{\dd\rho}{\rho}, \quad Q \equiv \frac{\dd P}{\rho},
\end{align}
the linearized equations for
locally isothermal dusty gas with the pressure equation in place of
the dust-fraction (Eq. \ref{masseq}---\ref{momeq}, Eq. \ref{eff_energy}) are then:

\begin{align}
  \ii\sigma W &= \ii k \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi + 
  \delta\bm{F}\cdot\hat{\bm{r}},\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &=  \delta\bm{F}\cdot\hat{\bm{z}},\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r \ln{c_s^2} - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy}
\end{align}  
where $^\prime \equiv \p_z$; $\dd\bm{F}$ is the linearized pressure force, given in Appendix
\ref{lin_press}; and 
%\begin{align}
%  \delta \bm{F} \equiv \frac{\dd\rho}{\rho^2}\nabla P -
%  \frac{1}{\rho}\nabla\dd P, 
%\end{align}
$\dd\mathcal{C}$ is the linearized dust-duffusion function, given in
Appendix \ref{lin_dust}.  We consider stopping times appropriate for
small grains in the Epstein regime. Note that for the axisymmetric
problem, $\dd\bm{F}$ is purely meridional.

Eq. \ref{lin_mass}---\ref{lin_energy} is a set of ordinary
differntial equations in $z$. All the coefficients are evaluated at a
fiducial radius $r=r_0$, but their full $z$-dependence is retained.  
\subsection{Boundary conditions}
Eq. \ref{lin_mass}---\ref{lin_energy} can be reduced to a set of
first-order differential equations for $W, Q$ and $\dd v_z$. We can
see this schematically as follows. Eq. \ref{lin_xmom} and
\ref{lin_ymom} may be combined to yield 
\begin{align*}
  \dd v_r = \dd v_r (W, Q, \dd v_z). 
\end{align*} 
We can then take the continuity equation as an equation for $\dd v_z$, 
\begin{align*}
\text{Eq. \ref{lin_mass}} \Rightarrow \dd v_z^\prime(z) = \dd
  v_z^\prime(W,Q,\dd v_z),
\end{align*}
and the vertical momentum equation as an equation for $Q$, 
\begin{align*}
\text{Eq. \ref{lin_zmom}} \Rightarrow Q^\prime(z) = 
 Q^\prime(W,Q,\dd v_z). 
\end{align*}

Now, for finite dust-gas coupling, $\tstop\neq0$, inspection of $\dd C$
(Appendix \ref{lin_dust}) shows that it involves $W^\prime$ (assuming
$\epsilon < 1$). Then Eq. \ref{lin_energy} may be taken as a
differential equation for $W$. However, for perfectly coupled dust,
$\tstop =\dd\mathcal{C}= 0$. In that case Eq. \ref{lin_energy} gives an algebraic
relation between $W$, $Q$ and $\dd v_z$. That is,
\begin{align*}
\text{Eq. \ref{lin_energy}}\Rightarrow \begin{cases}
  W^\prime(z) = W^\prime(W, Q, \dd v_z) & \text{if } \tstop \neq 0, \\
  W           (z) = W(Q, \dd v_z) & \text{if }\tstop = 0.
\end{cases}
\end{align*}

This means that for the perfectly coupled problem, we have a pair of
ODEs for $Q$ and $\dd v_z$, and two boundary conditions are needed. 
This is expected since this problem is exactly 
equivalent to that in adiabatic hydrodynamics \citep[e.g.]{lubow93}. When
$\tstop\neq0$, we require three boundary conditions. {\bf WHAT?? Why odd
  number of BCs? Shouldn't we have even number of BCs? We have
two boundaries? I've only seen problems like this involving even
number of BCs.} In this case we impose $\dd v_z =0$ at $z=\pm
z_\mathrm{max}$, and consider even modes about the midplane such that
$\dd v_z^\prime(0) = 0$.

>>>>

\section{Linear problem for perfectly coupled dust}\label{linear_problem}
In this paper we calculate explicit solutions to the linearized
equations in the limit of perfectly coupled dust ($\tstop=0$), so that
the equilibria defined in \S\ref{eqm} are exact steady states; and we
consider radially-localized axisymmetric disturbances of the form  
\begin{align}
  \delta f (r, z) = \delta f_1(r,z)\exp{(\ii k r)},
\end{align} 
%and similarly for $\dd P$ and $\dd\bm{v}$. 
where $k$ is a real wavenumber such that $|kr|\gg 1$, and the
amplitude $\dd f_1(r,z)$ is 
a slowly-varying function of $r$. Then 
$\p_r\to i k$ when acting on the above primitive perturbations, and we may
neglect curvature terms. We take  
$k>0$ without loss of generality. Hereafter, we drop the subscript 1
on the amplitudes. 
%The frequency $\sigma = \omega +
%\ii s$ is generally complex, with $\omega$ being the real frequency
%and $s$ is the real growth rate.

Introducing 
\begin{align}
  W \equiv \frac{\dd\rho}{\rho}, \quad Q \equiv \frac{\dd P}{\rho},
\end{align}
the linearized equations for 
locally isothermal, perfectly-coupled dusty gas with the pressure
equation in place of the dust-fraction
(Eq. \ref{masseq}---\ref{momeq}, Eq. \ref{eff_energy}) are then:

\begin{align}
  \ii\sigma W &= \ii k \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi + 
  \delta\bm{F}\cdot\hat{\bm{r}},\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &=  \delta\bm{F}\cdot\hat{\bm{z}},\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2}, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
%               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy} 
\end{align}  
where $^\prime \equiv \p_z$ and $\dd\bm{F}$ is the linearized pressure
force, given in Appendix \ref{lin_press}. Note that we have assumed a
temperature profile that only depends on $r$.

%; and 
%\begin{align}
%  \delta \bm{F} \equiv \frac{\dd\rho}{\rho^2}\nabla P -
%  \frac{1}{\rho}\nabla\dd P, 
%\end{align}
%$\dd\mathcal{C}$ is the linearized dust-duffusion function, given in
%Appendix \ref{lin_dust}. We consider stopping times appropriate for
%small grains in the Epstein regime. Note that for the axisymmetric
%problem, $\dd\bm{F}$ is purely meridional.

Eq. \ref{lin_mass}---\ref{lin_energy} is a set of ordinary
differntial equations in $z$. All coefficients and amplitudes are
evaluated at a fiducial radius $r=r_0$, but their full $z$-dependence
is retained.  Two boundary conditions are needed for the $\tstop=0$
problem. This is expected since this problem is exactly equivalent to
that in adiabatic hydrodynamics \citep[e.g.][]{lubow93}. For simplicity
we impose solid boundaries so that $\delta v_z(\pm\zmax)=0$ and
$\zmax=5\Hgas$.

We solve the linearized equations as a generalized eigenvalue problem 
using a pseudo-spectral code adapted from \cite{lin15}. Amplitudes 
are expanded in Chebyshev polynomials up to order $N_z-1$. Our 
standard resolution is $N_z=513$. We check results using
Eq. \ref{vsi_check}.

%\subsection{Boundary conditions}

%Eq. \ref{lin_mass}---\ref{lin_energy} can be reduced to a set of
%first-order differential equations for $W, Q$ and $\dd v_z$. We can
%see this schematically as follows. Eq. \ref{lin_xmom} and
%\ref{lin_ymom} may be combined to yield 
%\begin{align*}
%  \dd v_r = \dd v_r (W, Q, \dd v_z). 
%\end{align*} 
%We can then take the continuity equation as an equation for $\dd v_z$, 
%\begin{align*}
%\text{Eq. \ref{lin_mass}} \Rightarrow \dd v_z^\prime(z) = \dd
%  v_z^\prime(W,Q,\dd v_z),
%\end{align*}
%and the vertical momentum equation as an equation for $Q$, 
%\begin{align*}
%\text{Eq. \ref{lin_zmom}} \Rightarrow Q^\prime(z) = 
% Q^\prime(W,Q,\dd v_z). 
%\end{align*}

%Now, for finite dust-gas coupling, $\tstop\neq0$, inspection of $\dd C$
%(Appendix \ref{lin_dust}) shows that it involves $W^\prime$ (assuming
%$\epsilon < 1$). Then Eq. \ref{lin_energy} may be taken as a
%differential equation for $W$. However, for perfectly coupled dust,
%$\tstop =\dd\mathcal{C}= 0$. In that case Eq. \ref{lin_energy} gives an algebraic
%relation between $W$, $Q$ and $\dd v_z$. That is,
%\begin{align*}
%\text{Eq. \ref{lin_energy}}\Rightarrow \begin{cases}
%  W^\prime(z) = W^\prime(W, Q, \dd v_z) & \text{if } \tstop \neq 0, \\
%  W           (z) = W(Q, \dd v_z) & \text{if }\tstop = 0.
%\end{cases}
%\end{align*}

%This means that for the perfectly coupled problem, we have a pair of
%ODEs for $Q$ and $\dd v_z$, and 
%When
%$\tstop\neq0$, we require three boundary conditions. {\bf WHAT?? Why odd
%  number of BCs? Shouldn't we have even number of BCs? We have
%two boundaries? I've only seen problems like this involving even
%number of BCs.} In this case we impose $\dd v_z =0$ at $z=\pm
%z_\mathrm{max}$, and classify modes according to their symmetry about
%the midplane: `even' modes with $\dd v_z^\prime(0) = 0$, and `odd' 
%modes with $\dd v_z(0)=0$.

### 2016-07-30 model.tex
adding results

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding linearized equations about an
axisymmetric steady state. These lead to the integral relation,

\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\left|\nabla\cdot\left(P\dd
  \bm{v}\right)\right|^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV,\label{int_rel}
\end{align}
between the (complex) mode frequency $\sigma$ and the spatial
structure of the disturbances which are prefixed by deltas. 
The real coefficients $A,B,C,D$ can be read off
Eq. \ref{integral_ex}. Note that $B=C$ from dynamical equilibrium
(Eq. \ref{vshear}).

@@

\subsection{Locally isothermal gas perfectly coupled to dust} 
When $c_s(r,z)$ is non-uniform but $\tstop=0$, Eq. \ref{int_rel}
indicates that $\sigma^2$ is generally complex, implying instability.  
Writing $\sigma = 
\omega + \ii s$ where $\omega$ and $s$ are the real frequency and
growth rates, respectively, we have 
\begin{align}
  s = -\frac{\imag\int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}. 
\end{align} 
This is classic VSI caused by vertical shear arising from a radial
temperature gradient \citep{nelson13,barker15,lin15}.

@@

\subsection{Strictly isothermal gas well-coupled to dust} 
Now consider constant $c_s$ but $\tstop\neq0$. Then Eq. \ref{int_rel}
gives 
\begin{align}
  s = -\frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}, 
\end{align}
%In this case 
%the effective energy equation is 
%\begin{align}\label{dusty_cooling}
%  \frac{DP}{Dt} &= - P \nabla \cdot \bm{v} + \mathcal{C}
%\end{align} 
%The function $\mathcal{C}$ can be considered as a cooling term. 
%By
%analogy with enabling VSI in an adiabatic gas through cooling
%\citep{nelson13,lin15}, we may expect a `dusty VSI' to develop 
%if this term is large enough (physically meaning diffusion of the
%dust-fraction is strong). 
%Note that $\mathcal{C}$ is also responsible for the streaming
%instability  in the strong-drag limit \citep{laibe14}, though this
%does not require vertical stratification. In the form of
%Eq. \ref{dusty_cooling}, however, the streaming instability has a
%thermodynamical interpretation, similar to thermal instabilities in
%stars \citep{latter06}.  
which represent all instabilities due to large-but-finite dust-gas
friction. The formalism we have adopted give these instabilities a 
thermodynamical interpretation because the dust-gas friction enters the problem
as an energy source term \citep[cf. thermal instabilities in
  stars][]{latter06}. 
This includes the streaming instability in
the strong-drag limit \citep{youdin05a, jacquet11, laibe14}.

>>>>

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations about an axisymmetric steady state. A variable $f$ is
subject to perturbations of the form 
\begin{align}
  \delta f(r,z)\exp{\left(-\ii\sigma t\right)},
\end{align}
where $\sigma$ is the complex mode frequency. In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\left|\nabla\cdot\left(P\dd
  \bm{v}\right)\right|^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.\label{int_rel}
\end{align} 
The real coefficients $A,B,C,D$ can be read off 
Eq. \ref{integral_ex}. Note that $B=C$ from dynamical equilibrium
(Eq. \ref{vshear}). We now consider various limits of
Eq. \ref{int_rel}.

@@

\subsection{Locally isothermal gas perfectly coupled to dust}\label{dusty_vsi_int}
When $c_s(r,z)$ is non-uniform but $\tstop=0$, Eq. \ref{int_rel}
indicates that $\sigma^2$ is generally complex, implying instability.  
Writing $\sigma = 
-\omega + \ii s$ where $\omega$ and $s$ are the real frequency and
growth rates, respectively, we have 
\begin{align}
  s = \frac{\imag\int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}. \label{vsi_check}
\end{align} 
This is classic VSI caused by vertical shear arising from a radial
temperature gradient \citep{nelson13,barker15,lin15}. We compute
explicit solutions in this limit in
\S\ref{linear_problem}---\ref{results}.

@@

\subsection{Strictly isothermal gas well-coupled to dust} 
Now consider constant $c_s$ but $\tstop\neq0$. Then Eq. \ref{int_rel}
gives 
\begin{align}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}, 
\end{align}
%In this case 
%the effective energy equation is 
%\begin{align}\label{dusty_cooling}
%  \frac{DP}{Dt} &= - P \nabla \cdot \bm{v} + \mathcal{C}
%\end{align} 
%The function $\mathcal{C}$ can be considered as a cooling term. 
%By
%analogy with enabling VSI in an adiabatic gas through cooling
%\citep{nelson13,lin15}, we may expect a `dusty VSI' to develop 
%if this term is large enough (physically meaning diffusion of the
%dust-fraction is strong). 
%Note that $\mathcal{C}$ is also responsible for the streaming
%instability  in the strong-drag limit \citep{laibe14}, though this
%does not require vertical stratification. In the form of
%Eq. \ref{dusty_cooling}, however, the streaming instability has a
%thermodynamical interpretation, similar to thermal instabilities in
%stars \citep{latter06}.  
which represent all instabilities due to large-but-finite dust-gas
friction. The formalism we have adopted give these instabilities a 
thermodynamical interpretation because the dust-gas friction enters the problem
as an energy source term \citep[cf. thermal instabilities in
  stars][]{latter06}. 
This includes the streaming instability in
the strong-drag limit \citep{youdin05a, jacquet11, laibe14}.

### 2016-07-30 result.tex
adding results

\section{Results}

\begin{figure}
\includegraphics[width=\linewidth]{figures/eigenvalues}
\caption{do not believe}
\end{figure}

>>>>

\section{Perfectly coupled
  dusty vertical shear instability}\label{results}

When $\tstop=0$ we may construct exact dusty equilibra by specifying
the dust-to-gas ratio, as described in \S\ref{eqm}. We use this to
examine the effect of dust on classic VSI, defined to be driven by the
temperature gradient (\S\ref{dusty_vsi_int}). For simplicity we
consider constant input values of the midplane dust-to-gas ratio
$\tepsilon_0$ and the characteristic dust thickness $\Hd$. These,
together with the perturbation wavenumber $k$, comprises main the
parameters of the linear problem. We fix the radial power-law index
for the midplane gas density to $p = -1.5$, that for the
temperature profile to $q=-1$; and set the gas disk aspect-ratio
$h_\mathrm{g}=0.05$. These are fiducial values used in \citetalias{lin15}.

\subsection{Effect of dust-loading}
We first vary the dust-to-gas ratio $\tepsilon_0\in[10^{-3},1]$,
fixing the dust thickness to $\Hd=0.99\Hg$ (so that $\tepsilon$ is
roughly constant with height) and perturbation
wavenumber to $k\Hg = 30$.

Fig. \ref{vsi_dust_loading} show unstable
modes found for different values of $\tepsilon_0$. The eigenvalue
distributions for $\tepsilon_0 \leq 10^{-2}$ are similar to the
dust-free fiducial case considered by \citetalias{lin15}, consisting
of the roughly horizontal `body modes' and the nearly-vertical
`surface modes'. The latter is due to the imposed vertical boundaries
\citep{barker15}.

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably `surface modes', which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\tepsilon_0\geq
0.1$. The body modes' growth rates remain $\sim O(h\OmK)$ but their
oscillation frequency increases with $\tepsilon_0$.

In Fig. \ref{vsi_dust_loading2d} we compare the

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_kx30Hd1} 
  \caption{Unstable modes in a locally isothermal, perfectly coupled
    dusty disk with fiducial parameters
    $(p,q,h_\mathrm{g}, \Hd/\Hg )=(-1.5,-1,0.05, 0.99)$. The real
    frequency $\omega$ and growth rates $s$ are shown for a range of
    midplane dust-to-gas ratios $\tepsilon_0=\rhod/\rhog$. 
    \label{vsi_dust_loading}
    }
\end{figure}

\begin{figure}
  \includegraphics[scale=0.54, clip=true, trim=0cm 2.5cm 0cm 0cm]{figures/result2d_dg1d-3.ps}\\
  \includegraphics[scale=0.54]{figures/result2d_dg1.ps} 
  \caption{Fundamental dusty VSI mode in real space for midplane dust-to-gas
    ratio $\tepsilon_0=10^{-3}$ (top) and $\tepsilon_0=1$
    (bottom). The color scale shows the perturbation to the
    dust-to-gas ratio, $\delta\tepsilon$; and the arrows show
    $\sqrt{\rho}\left(\dd v_x, \dd v_z\right)$. 
    \label{vsi_dust_loading2d}
    }
\end{figure}

\subsection{Effect of dust thickness}
